An Efficient Minibatch Acceptance Test for Metropolis-Hastings

### 2016-07-06 nips_2016.tex
updates to first section

\title{An Efficient Minibatch Acceptance Test for MCMC}

@@

\begin{abstract}
Markov chain Monte Carlo (MCMC) methods have many applications in
machine learning. We are particularly interested in their application
to modeling very large datasets, where it is impractical to
perform Metropolis-Hastings tests on the full data. Previous work on reducing
the cost of Metropolis-Hastings tests yield variable data reductions
and in the worst case require the entire dataset. 
Here we present a simple test which requires constant size minibatches
if an auxiliary condition is met. We describe several cases under which
this condition can be satisfied. Our new test is naturally tolerant
of ``noisy'' estimates of the transition probability ratio. To achieve
this, we replace the classical Metropolis-Hastings test with a smooth
acceptance function which is a sigmoid (a logistic function). 
The resulting test can be combined with minibatch proposals to yield
updates with the same complexity as a simple SGD update. In this paper we
derive the test, discuss its implementation, and present several
experiments.
\end{abstract}

@@

Consider a simple model where samples $x_i$ for $i=1,\ldots,N$ are
drawn from a Normal distribution $\mathcal{N}(\theta,1)$ where
$\theta=0.5$, and with a uniform prior on $\theta$. The log likelihood ratio is
\begin{equation}
\Lambda^*(\theta,\theta') =  \frac{1}{b}\sum_{i=1}^b \log\frac{p(x_i|\theta')}{p(x_i|\theta)}=
  %%\frac{1}{2}\left( (x_i-\theta')^2 - (x_i - \theta)^2\right) =
  (\theta'-\theta)\left(\frac{1}{b}\sum_{i=1}^b x_i-\theta-\frac{\theta'-\theta}{2}\right)
\end{equation}
which is normally distributed over selection of the Normal samples $x_i$.
Since the $x_i$ have unit variance, their mean has variance $1/b$, and the variance $\sigma^2(\Lambda^*)$
of $\Lambda^*(\theta,\theta')$ is $(\theta'-\theta)^2/b$. In order to pass
a hypothesis test that $\Lambda > \psi$, we will need to establish a
large enough gap (several $\sigma(\Lambda^*)$) between $\Lambda^*(\theta',\theta)$
and $\psi(u,\theta',\theta)$.

@@

\begin{lemma}\label{lem:efficienty_cutting}
  For the model above, there
  exists a fixed (independent of $N$) constant $c$ such that with probability $ \geq c$ over the joint
  distribution of $(\theta, \theta', u)$, the
  test from \cite{icml2014c1_bardenet14} consumes all $N$ samples. 
\end{lemma}

@@

  The samples $\theta$, $(\theta'-\theta)$
  and $u$ are drawn independently and so the probability of the conjunction of these events is
  $c = p_0 p_1 p_2 = 0.009$.
\end{proof}
There is a separate line of MCMC work drawing principles from
statistical physics. By viewing random variables as particles in a
system, one can apply Hamiltonian Monte Carlo
(HMC)~\cite{mcmc_hamiltonian_2010} methods which generate high
acceptance \emph{and} distant proposals when run on full batches of
data. Recently Langevin
Dynamics~\cite{langevin_2011,conf/icml/AhnBW12} has been applied to
Bayesian estimation on minibatches of data. This simplified dynamics
uses local proposals and avoids MH tests by using small proposal steps
whose acceptance approaches 1 in the limit. However, the constraint on
proposal step size is severe, and the state space exploration reduces
to a random walk. Full minibatch HMC for minibatches was recently
described in ~\cite{sghmc_2014} which allows momentum-augmented
proposals with larger step sizes. However, step sizes are still
limited by the need to run accurately without MH tests.
% \cite{stochastic_thermostats_2014}
Our work opens the door to applying those methods with much
more aggressive step sizes without loss of accuracy. We demonstrate
this in Section~\ref{ssec:nets}.

>>>>

\title{An Efficient Minibatch Acceptance Test for Metropolis-Hastings}

@@

\begin{abstract}
Markov chain Monte Carlo (MCMC) methods have many applications in
machine learning. We are particularly interested in their application
to modeling very large datasets, where it is impractical to perform
Metropolis-Hastings tests on the full data. Previous work on reducing
the cost of Metropolis-Hastings tests yield variable data reductions
and in the worst case require the entire dataset.  Here we present a
simple test which requires constant size minibatches if an auxiliary
condition is met. We describe several cases under which this condition
can be satisfied. Our new test is naturally tolerant of ``noisy''
estimates of the transition probability ratio. To achieve this, we
replace the classical Metropolis-Hastings acceptance test with a
smooth acceptance function which is a sigmoid (logistic)
function of the log likelihood ratio.
The resulting test can be combined with minibatch proposals to yield
updates with the same complexity as a simple SGD update. In this paper
we derive the test, discuss its implementation, and present several
experiments.
\end{abstract}

@@

\subsection{An Example}
Consider a simple model where samples $x_i$ for $i=1,\ldots,N$ are
drawn from a Normal distribution $\mathcal{N}(\theta,1)$ where
$\theta=0.5$, and with a uniform prior on $\theta$. The log likelihood ratio is
\begin{equation}
\Lambda^*(\theta,\theta') =  \frac{1}{b}\sum_{i=1}^b \log\frac{p(x_i|\theta')}{p(x_i|\theta)}=
  %%\frac{1}{2}\left( (x_i-\theta')^2 - (x_i - \theta)^2\right) =
  (\theta'-\theta)\left(\frac{1}{b}\sum_{i=1}^b x_i-\theta-\frac{\theta'-\theta}{2}\right)
\end{equation}
which is normally distributed over selection of the Normal samples $x_i$.
Since the $x_i$ have unit variance, their mean has variance $1/b$, and the variance $\sigma^2(\Lambda^*)$
of $\Lambda^*(\theta,\theta')$ is $(\theta'-\theta)^2/b$. In order to pass
a hypothesis test that $\Lambda > \psi$, we will need to establish a
large enough gap (several $\sigma(\Lambda^*)$) between $\Lambda^*(\theta',\theta)$
and $\psi(u,\theta',\theta)$.

@@

\begin{lemma}\label{lem:efficienty_cutting}
  For the model above, there
  exists a fixed (independent of $N$) constant $c$ such that with probability $ \geq c$ over the joint
  distribution of $(\theta, \theta', u)$, the
  tests from \cite{icml2014c1_bardenet14} and \cite{cutting_mh_2014} consume all $N$ samples. 
\end{lemma}

@@

  The samples $\theta$, $(\theta'-\theta)$
  and $u$ are drawn independently and so the probability of the conjunction of these events is
  $c = p_0 p_1 p_2 = 0.009$.
\end{proof}

Similar results can be proved for other distributions and proposals, by identifying
regions in product space $(\theta,\theta'-\theta,u)$ such that the hypothesis test
needs to separate nearly-equal values.

It follows that the accelerated M-H tests in \cite{icml2014c1_bardenet14} and
\cite{cutting_mh_2014} require at least a constant fraction $\geq c$
in the amount of data consumed per test, compared to full-dataset tests.
i.e. their speed-up is at most $1/c$.

\subsection{MCMC Posterior Inference}
There is a separate line of MCMC work drawing principles from
statistical physics. By viewing random variables as particles in a
system, one can apply Hamiltonian Monte Carlo
(HMC)~\cite{mcmc_hamiltonian_2010} methods which generate high
acceptance \emph{and} distant proposals when run on full batches of
data. Recently Langevin
Dynamics~\cite{langevin_2011,conf/icml/AhnBW12} has been applied to
Bayesian estimation on minibatches of data. This simplified dynamics
uses local proposals and avoids MH tests by using small proposal steps
whose acceptance approaches 1 in the limit. However, the constraint on
proposal step size is severe, and the state space exploration reduces
to a random walk. Full minibatch HMC for minibatches was recently
described in ~\cite{sghmc_2014} which allows momentum-augmented
proposals with larger step sizes. However, step sizes are still
limited by the need to run accurately without MH tests.
% \cite{stochastic_thermostats_2014}
Our work opens the door to applying those methods with much
more aggressive step sizes without loss of accuracy. We demonstrate
this in Section~\ref{ssec:nets}.
